---
title: "1_datawrangle"
output: html_document
date: "2022-09-27"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

**This markdown goes through the initial filtering stages for Pocillopora meandrina using Dartseq data** 

#Load libraries
```{r}
library(dartR)
library(adegenet)
library(poppr)
library(hierfstat)
library(reshape2)
library(ggplot2)
library(RColorBrewer)
library(scales)
library(dplyr)
```


```{r}
gl <- gl.read.dart(filename = "Report_DPoc22-7167_SNP_mapping_2.csv", covfilename = "Pmeandrina_dartLIMS_sector.csv")
#saveRDS(gl, file="gl.Rdata") 

gl <- gl.drop.ind(gl,
   ind.list=c('JAN2019_13-124_P. verrucosa_103','P ver 221', 'P ver 371', 'P ver 372', 'P ver 291'))

#we know that 62 total samples failed, so this makes sense that we only have 115 / 188 individuals recognised in our data frame. 
```

##Checking metadata 
```{r}
nInd(gl)
#115 individuals

nLoc(gl)
#43980 loci in the original dataset (gl)

nPop(gl)
#13 populations ; regroup GBR by sub-regions 
popNames(gl)


#converting SNP files to a matrix 
#generate a matrix of the SNP scores, with 0 as homozygous reference, 2 as homozygous alternate and 1 as heterozygous
m <- as.matrix(gl)
```


```{r}
#Dropping individiuals with weird admixture to see if it improves the number of loci retained
#JAN2019_13-124_P. verrucosa_103
#P ver 221
#P ver 371
#P ver 372
#P ver 291

gl.drop <- gl.drop.ind(gl,
   ind.list=c('JAN2019_13-124_P. verrucosa_103','P ver 221', 'P ver 371', 'P ver 372', 'P ver 291'))

nInd(gl.drop)
#110 ind now

nLoc(gl.drop)
#same loci currently. See if next parameters change 
```


##Filter by call rate
```{r}
#First we are filtering SNP call rate typically at 95 % threshold to make a confident call on genotype. 
#we need to decide whether it's better to filter by individual or by locus first. We agree it is better to lose loci in this case than to lose individuals, so we will filter by loci first. Indicated by method = "loc"
# We are basically calculating the percentage of complete genotypes per loci in the entire dataset for Acropora. 

#filter out secondaries (all but one SNP retained per sequence tag)
gl <- gl.filter.secondaries(gl)

nLoc(gl)
#now we have 27,253 loci 

gl.report.reproducibility(gl)

#filter loci with reproducibility below a particular threshold
gl <- gl.filter.reproducibility(gl, threshold = 0.98)

nLoc(gl)
#22,055

#filter loci with >5% missing data. this is standard. 
gl2 <- gl.filter.callrate(gl, method = "loc", threshold = 0.85)
nLoc(gl2)
#1827 SNPs remaining 

#provide a summary of the call rate values
gl.report.callrate(gl2) 
```



##Filter by call rate
```{r}
#First we are filtering SNP call rate typically at 95 % threshold to make a confident call on genotype. 
#we need to decide whether it's better to filter by individual or by locus first. We agree it is better to lose loci in this case than to lose individuals, so we will filter by loci first. Indicated by method = "loc"
# We are basically calculating the percentage of complete genotypes per loci in the entire dataset for Acropora. 

#filter out secondaries (all but one SNP retained per sequence tag)
gl.drop <- gl.filter.secondaries(gl.drop)

nLoc(gl.drop)
#same : now we have 27,253 loci 

gl.report.reproducibility(gl.drop)

#filter loci with reproducibility below a particular threshold
gl.drop <- gl.filter.reproducibility(gl.drop, threshold = 0.98)

nLoc(gl.drop)
#a few less: 22000

#filter loci with >5% missing data. this is standard. 
gl2.drop <- gl.filter.callrate(gl.drop, method = "loc", threshold = 0.85)
nLoc(gl2.drop)
#1856 SNPs remaining ; slightly more with dropped 5 individuals 

#provide a summary of the call rate values
gl.report.callrate(gl2.drop) 
```


## MAF filtering 
```{r}
#We will filter out loci with a MAF less than 0.05.**
gl3 <- gl.filter.maf(gl2, threshold = 0.95) 


#report for MAF after filtering with MAF < 0.05 
gl.report.maf(gl3)
nLoc(gl3)
#1260 snps remaining 
```

## MAF filtering 
```{r}
#We will filter out loci with a MAF less than 0.05.**
gl3.drop <- gl.filter.maf(gl2.drop, threshold = 0.95) 


#report for MAF after filtering with MAF < 0.05 
gl.report.maf(gl3.drop)
nLoc(gl3.drop)
#1117 snps remaining 
```

##Impute missing data 
```{r, include = FALSE}
#here we are imputing any missing data using the 'neighbour' method. this is essential for some analyses later on to not have missing data 
gl3 <- gl.impute(
  gl3,
  method = "neighbour",
  fill.residual = TRUE,
  parallel = FALSE,
  verbose = NULL
)
```



##Converting to a genlight object for downstream analysis 
```{r}
#converting a genlight object (we have) to a genind object so we can look at genetic data at an individual data using adegenet package. 
pmea_gen <- gl2gi(gl3, probar = FALSE, verbose = NULL)
pmea_gen

is.genind(pmea_gen)
```


## Checking relatedness (for clones)
```{r}
#here we are checking for any clones / duplicate samples. We want to remove any that are. 
#check genotypes are unique
mlg(pmea_gen)

#so we don't have any clones 
```


```{r}
#check that loci are still polymorphic after filtering. Retaining monomorphic loci unnecessarily increases the size of the dataset and will affect some calculations.

isPoly(pmea_gen) %>% summary

#remove loci that are not polymorphic 

poly_loci = names(which(isPoly(pmea_gen) == TRUE))
pmea_gen = pmea_gen[loc = poly_loci]
isPoly(pmea_gen) %>% summary
```


```{r}
#Here we are calculating basic stats using hierfstat
basic_pmea = basic.stats(pmea_gen, diploid = TRUE)

# Mean observed heterozygosity per site
Ho_pmea = apply(basic_pmea$Ho, MARGIN = 2, FUN = mean, na.rm = TRUE) %>%
  round(digits = 2)
Ho_pmea

# Mean expected heterozygosity per site
He_pmea = apply(basic_pmea$Hs, MARGIN = 2, FUN = mean, na.rm = TRUE) %>%
  round(digits = 2)
He_pmea

# Create a data.frame of site names, Ho and He and then convert to long format
Het_pmea_df = data.frame(Site = names(Ho_pmea), Ho = Ho_pmea, He = He_pmea) %>%
  melt(id.vars = "Site")

# Custom theme for ggplot2
custom_theme = theme(
  axis.text.x = element_text(size = 10, angle = 90, vjust = 0.5, face = "bold"),
  axis.text.y = element_text(size = 10),
  axis.title.y = element_text(size = 12),
  axis.title.x = element_blank(),
  axis.line.y = element_line(size = 0.5),
  legend.title = element_blank(),
  legend.text = element_text(size = 12),
  panel.grid = element_blank(),
  panel.background = element_blank(),
  plot.title = element_text(hjust = 0.5, size = 15, face="bold")
  )

# Italic label
hetlab.o = expression(italic("H")[o])
hetlab.e = expression(italic("H")[e])

# Acropora heterozygosity barplot
ggplot(data = Het_pmea_df, aes(x = Site, y = value, fill = variable))+
  geom_bar(stat = "identity", position = position_dodge(width = 0.6), colour = "black")+
  scale_y_continuous(expand = c(0,0), limits = c(0,0.3), breaks = c(0, 0.05, 0.10, 0.15, 0.20, 0.25))+
  scale_fill_manual(values = c("pink", "#bdbdbd"), labels = c(hetlab.o, hetlab.e))+
  ylab("Heterozygosity")+
  ggtitle("P. meandrina")+
  custom_theme
```


```{r}
#Looking at Fis values 
apply(basic_pmea$Fis, MARGIN = 2, FUN = mean, na.rm = TRUE) %>%
  round(digits = 3)
```

## Calculate Fst values 
```{r}
#using hierfstat package here. 
#compute pairwise Fst values. Returns pairwise distances folloiwng Weir & Cockerham 1984 
pmea_fst = genet.dist(pmea_gen, method = "WC84") %>% round(digits = 3)
pmea_fst
```

##Plotting Fst values as a grid 
```{r}
library(wesanderson)
pal = wes_palette("Zissou1", 12, type = "continuous")

#Visualise pairwise Fst for reefs 
# Desired order of labels
lab_order = c("Wreck", "Saumarez", "Frederick", "Marion", "Flinders", "Chilcott", "Holmes", "Moore", "Bougainville", "Osprey", "13-124", "21-550", "22-084", "Corbett", "Chinaman", "Chicken", "Davie", "Davies", "Fitzroy", "Hoskyns", "Kelso", "Lady Musgrave")

# Change order of rows and cols
fst.mat = as.matrix(pmea_fst)
fst.mat1 = fst.mat[lab_order, ]
fst.mat2 = fst.mat1[, lab_order]

# Create a data.frame
ind = which(upper.tri(fst.mat2), arr.ind = TRUE)
fst.df = data.frame(Site1 = dimnames(fst.mat2)[[2]][ind[,2]],
                    Site2 = dimnames(fst.mat2)[[1]][ind[,1]],
                    Fst = fst.mat2[ ind ])

# Keep the order of the levels in the data.frame for plotting 
fst.df$Site1 = factor(fst.df$Site1, levels = unique(fst.df$Site1))
fst.df$Site2 = factor(fst.df$Site2, levels = unique(fst.df$Site2))

# Convert minus values to zero
fst.df$Fst[fst.df$Fst < 0] = 0

# Print data.frame summary
fst.df %>% str


# Fst italic label
fst.label = expression(italic("F")[ST])

# Extract middle Fst value for gradient argument
mid = max(fst.df$Fst) / 2

# Plot heatmap
ggplot(data = fst.df, aes(x = Site1, y = Site2, fill = Fst))+
  geom_tile(colour = "black")+
  geom_text(aes(label = Fst), color="black", size = 3)+
    scale_fill_gradientn(colours = pal) + 
  scale_colour_gradientn(colours = pal) +
  scale_x_discrete(expand = c(0,0))+
  scale_y_discrete(expand = c(0,0), position = "right")+
  theme(axis.text = element_text(colour = "black", size = 10, face = "bold"),
        axis.title = element_blank(),
        axis.text.x = element_text(angle = 45, vjust = 0.5), 
        panel.grid = element_blank(),
        panel.background = element_blank(),
        legend.position = "right",
        legend.title = element_text(size = 14, face = "bold"),
        legend.text = element_text(size = 10)
        )
```



#Admixture plots 
```{r}
library(LEA)
## Here we are moving on to the LEA package in order to estimate K clusters, create admixture plots, impute missing data
```


```{r}
#convert genlight object to a geno object for LEA
gl2geno(gl3, outfile = "gl_geno", outpath = getwd(), verbose = NULL)
```

```{r}
pc = pca("gl_geno.lfmm", scale = TRUE)
tw = tracy.widom(pc)
```


```{r}
#Plots the percentage of variance explained by eah component. You can look at the 'elbow' to inform the number of genetic clusers. It looks like ours is between 3-5.
plot(tw$percentage, pch = 19, col = "darkblue", cex = .8)
```

```{r}
project = NULL
project = snmf("gl_geno.geno",
K = 1:10,
entropy = TRUE,
repetitions = 10,
project = "new")
```

```{r}
# plot cross-entropy criterion for all runs in the snmf project
plot(project, col = "blue", pch = 19, cex = 1.2)
```


```{r}
# select the best run for K = 4 clusters
best = which.min(cross.entropy(project, K = 3))
my.colors <- c("tomato", "lightblue",
"olivedrab", "gold")
barchart(project, K = 3, run = best,
border = NA, space = 0,
col = my.colors,
xlab = "Individuals",
ylab = "Ancestry proportions",
main = "Ancestry matrix") -> bp
axis(1, at = 1:length(bp$order),
labels = bp$order, las=1,
cex.axis = .4)
```



```{r}
pmea_gen_all <- gl2gi(gl3, probar = FALSE, verbose = NULL)
pmea_gen_all

is.genind(pmea_gen_all)
```


##Running PCA before DAPC
```{r}
x = tab(pmea_gen_all, NA.method = "mean")

# Perform PCA
pca1 = dudi.pca(x, scannf = FALSE, scale = FALSE, nf = 3)

# Analyse how much percent of genetic variance is explained by each axis
percent = pca1$eig/sum(pca1$eig)*100

barplot(percent, ylab = "Genetic variance explained by eigenvectors (%)", ylim = c(0,12),
        names.arg = round(percent, 1))
```


##Visualise PCA 
```{r}
colourCount = length(unique(pmea_gen_all$pop))
getPalette = colorRampPalette(brewer.pal(9, "Set1"))

# Create a data.frame containing individual coordinates
ind_coords = as.data.frame(pca1$li)

# Rename columns of dataframe
colnames(ind_coords) = c("Axis1","Axis2","Axis3")

# Add a column containing individuals
ind_coords$Ind = indNames(pmea_gen_all)

# Add a column with the site IDs
ind_coords$Site = pmea_gen_all$pop

# Calculate centroid (average) position for each population
centroid = aggregate(cbind(Axis1, Axis2, Axis3) ~ Site, data = ind_coords, FUN = mean)

# Add centroid coordinates to ind_coords dataframe
ind_coords = left_join(ind_coords, centroid, by = "Site", suffix = c("",".cen"))

# Define colour palette
#cols = brewer.pal(nPop(pmea_gen_all), "Set3")

# Custom x and y labels
xlab = paste("Axis 1 (", format(round(percent[1], 1), nsmall=1)," %)", sep="")
ylab = paste("Axis 2 (", format(round(percent[2], 1), nsmall=1)," %)", sep="")

# Custom theme for ggplot2
ggtheme = theme(axis.text.y = element_text(colour="black", size=12),
                axis.text.x = element_text(colour="black", size=12),
                axis.title = element_text(colour="black", size=12),
                panel.border = element_rect(colour="black", fill=NA, size=1),
                panel.background = element_blank(),
                plot.title = element_text(hjust=0.5, size=15) 
)

# Scatter plot axis 1 vs. 2
ggplot(data = ind_coords, aes(x = Axis1, y = Axis2))+
  geom_hline(yintercept = 0)+
  geom_vline(xintercept = 0)+
  # spider segments
  geom_segment(aes(xend = Axis1.cen, yend = Axis2.cen, colour = Site), show.legend = FALSE)+
  # points
  geom_point(aes(fill = Site), shape = 21, size = 3, show.legend = FALSE)+
  # centroids
  geom_label(data = centroid, aes(label = Site, fill = Site), size = 4, show.legend = FALSE)+
  # colouring
     scale_fill_manual(values = getPalette(colourCount))+
  scale_colour_manual(values = getPalette(colourCount))+
  # custom labels
  labs(x = xlab, y = ylab)+
  ggtitle("P mea All PCA")+
  # custom theme
  ggtheme

```


#Admixture plots 
```{r}
library(LEA)
## Here we are moving on to the LEA package in order to estimate K clusters, create admixture plots, impute missing data
```

```{r}
#convert genlight object to a geno object for LEA
gl2geno(gl3, outfile = "gl_geno", outpath = getwd(), verbose = NULL)
```

```{r}
pc = pca("gl_geno.lfmm", scale = TRUE)
tw = tracy.widom(pc)
```

```{r}
#Plots the percentage of variance explained by eah component. You can look at the 'elbow' to inform the number of genetic clusers. It looks like ours is between 3-5.
plot(tw$percentage, pch = 19, col = "darkblue", cex = .8)
```


#Run snmf algorithm
```{r, include=FALSE}
snmf1 = snmf("gl_geno.geno",
        K = 1:10,    #number of K ancestral populations to run
        entropy = TRUE,     #calculate cross-entropy
        repetitions = 10,    #10 repetitions for each K
        project = "new")
```

```{r}
plot(snmf1, col = "blue", cex = 1.5, pch = 19)
```


```{r}
# Extract the cross-entropy of all runs where K = 3
ce = cross.entropy(snmf1, K = 3)
ce
```

```{r}
lowest.ce = which.min(ce)
lowest.ce
```

```{r}
qmatrix = as.data.frame(Q(snmf1, K = 3, run = lowest.ce))
head(qmatrix)
```

```{r}
# Label column names of qmatrix
ncol(qmatrix)
cluster_names = c()
for (i in 1:ncol(qmatrix)){
  cluster_names[i] = paste("Cluster", i)
}
cluster_names
colnames(qmatrix) = cluster_names
head(qmatrix)

```

```{r}
# Add individual IDs
qmatrix$Ind = indNames(pmea_gen)
```

```{r}
#Add site IDs
qmatrix$Site = pmea_gen$pop
head(qmatrix)
```

```{r}
#Convert dataframe to long format
qlong = melt(qmatrix, id.vars=c("Ind","Site"))
head(qlong)
```

```{r}
#Adjust facet labels
levels(qlong$Site)
# Change order of sites by using the factor function
site.order = c("PCB", "Townsville", "Swains","CapBunk", "Bougainville", "Moore", "Willis", "Holmes", "Chilcott", "Marion", "Frederick", "Saumarez", "Wreck")
qlong$Site_ord = factor(qlong$Site, levels = site.order)

qlong <- qlong %>% mutate(site.order = factor(Site, levels = c("PCB",  "Townsville", "Swains", "CapBunk","Bougainville", "Moore", "Willis", "Holmes", "Chilcott", "Marion", "Frederick", "Saumarez", "Wreck")))
levels(qlong$Site)

#define colour palette 
pal = colorRampPalette(c("tomato", "lightblue",
"olivedrab", "gold"))
cols = pal(length(unique(qlong$variable)))

admix.bar = ggplot(data=qlong, aes(x=Ind, y=value, fill=variable))+
  geom_bar(stat = "identity")+
  scale_y_continuous(expand = c(0,0))+
  facet_wrap(~site.order, scales = "free", ncol = 4)+
  scale_fill_manual(values = cols)+
  ylab("Admixture proportion")+
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        axis.title.x = element_blank(),
        strip.text = element_text(colour="black", size=12),
        panel.grid = element_blank(),
        panel.background = element_blank(),
        legend.position = "top",
        legend.title = element_blank(),
        legend.text = element_text(size = 12))
admix.bar
```

```{r}
#Which individuals are only in cluster 2? 
qlong %>% filter(variable == "Cluster 2") %>% 
  filter(value > 0.75)

#Let's try filtering out these individuals and see if we retain more loci
#JAN2019_13-124_P. verrucosa_103
#P ver 221
#P ver 371
#P ver 372
#P ver 291
```

