---
title: "2_GEAv3"
author: "M.Marzonie"
date: "2023-12-14"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#Load libraries 
```{r}
library(dartR)
library(adegenet)
library(poppr)
library(hierfstat)
library(reshape2)
library(ggplot2)
library(RColorBrewer)
library(scales)
library(dplyr)
library(adegenet)
```

#1. Genetic data
##1.1 Calculate individual genetic distance
```{r}
#genind format genetic matrix 
load("acro_gen.RData")
gen <- acro_gen %>% as.matrix


dim(gen)
#260 individuals and 8332 loci 
sum(is.na(gen))
#0 because we already imputed missing data. 
```

```{r}
gen.imp <- apply(gen, 2, function(x) replace(x, is.na(x), as.numeric(names(which.max(table(x))))))
sum(is.na(gen.imp)) # No NAs
```


#2. Spatial data 
##2.1 Calculate Spatial data (MEMs)
```{r}
#Moran Eigenvector Maps for spatial environment 
library('codep')
library('adespatial')
library('rnaturalearth')
library('rnaturalearthdata')
library('raster')

geo = read.csv('ahum.recifs.csv') %>% 
  rename(pop = ID) %>% 
  mutate(ID = 1:nrow(.)) %>% 
  mutate(ID = as.character(ID))
              
# calculate distance between sampling locations
DistSpatial=gcd.hf(geo[,2:3]) 

# compute dbMEMs
dbmem = dbmem(DistSpatial)

# get countries polygons for plotting
landPolygons = ne_countries(scale = 'medium')

# create colorpalette for MEM
paletteMEM = colorRampPalette(c('red','yellow','green2'))(20)

# select MEM of interest
par(mfrow=c(2,3))

for (i in 1:3) {
  
MoI = paste0('MEM',i)
plot(geo[,3:2], pch=21, bg=paletteMEM[cut(dbmem[,MoI], breaks = 20)])
plot(landPolygons, add=T, col='grey')

}

```


##2.2 Extract dbMEMs
```{r}
# Convert dbmem to a data frame
dbmem_df <- as.data.frame(dbmem) %>% 
  tibble::rownames_to_column(var = "ID")

# Left join 'dbmem_df' with the 'geo' data frame based on the 'ID' column
dbmem_with_reef <- dbmem_df %>%  left_join(geo)

dbmem_pca <- dbmem_with_reef %>% dplyr::select(c(pop, MEM1, MEM2, MEM3)) %>% 
  rename(Reef = pop)

#save(dbmem_pca, file = "dbmem_pca.RData")
```


#3 Non-thermal environmental data
##3.1 Calculate Env PC1 and PC2
```{r}
library(tidyverse)
library(vegan)
library(GGally)
library(corrplot)
library(car)
library(scales)
library(ggrepel)
library(gclus)
library(ape)
#set up non-thermal variables. Remove any DHW-related variables. 
# i think will need to set up a dataframe with environmental variables only? Unsure. 
acro.recifs <- read.csv("Ahum_nonthermal.csv") %>% 
  dplyr::select(-ID, -POI_LON, -POI_LAT, -DIST)


nontherm.var <- acro.recifs %>% dplyr::select(POI_ID, CHL_me_002, FE_me_002, O2_me_002, PH_me_002, NO3_me_002, PO4_me_002, SPM_me_002, SSS_me_002, SCV_me_002) %>% 
      rename(Reef = POI_ID) 


#save(nontherm.var, file = "nontherm.var.RData")

#scale and centre the data but not lat/lon
envdat.phy.scaled <- scale(nontherm.var[,-c(1:3)],
                          center=TRUE, # center data by subtracting column means 
                           scale=TRUE) # divide centered data by standard deviation

#run RDA function for PCA without the ID/lat/long
env.pca <- rda(envdat.phy.scaled[,-c(1:3)])  

#run the PCA 
env.pca

summary(env.pca, display = NULL)
screeplot(env.pca)

#extract first two PCs
selected_components <- scores(env.pca, choices=1:2, display='sites')

# Combine the selected principal components with the POI_ID (reef names)
proxy_data <- cbind(acro.recifs["POI_ID"], selected_components)

envPC_data <- proxy_data %>% 
  rename(Reef = POI_ID) %>% 
  rename(Env.PC1 = PC1) %>% 
  rename(Env.PC2 = PC2) 
# Print the first few rows of the proxy_data dataframe to verify the results
head(envPC_data)

```

#4. Host Cluster data 
##4.1 Load host cluster admixture 
```{r}
load("qmatrix.RData")
hostcluster <- qmatrix %>% 
  rename(Reef = Site)

```

#5 Join data sets 

```{r}
#join cluster + non-thermal environment 
host.env <- hostcluster %>% left_join(., envPC_data)

#join above with spatial data and make ind and reef factors 
all.var <- host.env %>% left_join(., dbmem_pca) %>% 
  mutate(Ind = as.factor(Ind)) %>% 
  mutate(Reef = as.factor(Reef))

#changing ind and reef to be at the beginning of the dataframe 
all.var <- all.var %>% dplyr::select(Reef, everything()) %>% 
dplyr::select(Ind, everything()) 


library(stringr)

all.var <- all.var %>%
  mutate(Ind = if_else(str_detect(Ind, 'A ten193'), 'A ten 193', Ind))


load("snp.acro.RData")
acro.depth <- snp.acro %>% dplyr::select(Vial, Depth) %>% 
  rename(Ind = Vial)

all.var <- all.var %>% left_join(acro.depth) 
  
```

#6. Full RDA
##6.1 
```{r}
pred <- all.var[3:11]
```

```{r}
pred.pca <- rda(pred, scale = T)
summary(pred.pca)$cont
```

```{r}
screeplot(pred.pca, main = "Screeplot: Eigenvalues of Acropora Predictor Variables")
```

```{r}
round(scores(pred.pca, choices=1:8, display="species", scaling=0), digits=3)
```
**We know that 40% of variance in predictors is explaiend by the first PC axis and 28% by second PC axis. The strongest predictors associated with PC1 are recent meanDHW, DHW2020, meanSST, and maxSST, and MMM**


```{r}
acro.rda <- rda(gen.imp ~ ., data = pred, scale=T)
acro.rda
```

##Run backward stepwise model
```{r}
library(vegan)
# Use backward selection to identify significant thermal environmental variables
ord_acro <- dbrda(gen ~ ., data = pred)

sort(vif.cca(ord_acro)) 

os_acro_back <- ordistep(ord_acro, direction = "both", permutations = 999)
anova(os_acro_back, by = 'margin')

```

```{r}
RsquareAdj(acro.rda)
#adjusted R squared does not explain a lot. we need to use conditional factors to explain population structure before running the thermal models. 
#environmental (non thermal) variables and host cluster accounts for 5.7% of variation of neutral genetic structure 
```

**So we will include significant environmental factors as conditional effects in the thermal model. Cluster 2, Cluster 3, EnvPC1, EnvPC2, MEM1, MEM2, and MEM3**

#7. Run partial RDA
##7.1 Join thermal metrics to dataset 
```{r}
load("site.therm.RData")
load("snp.acro.RData")
therm <- snp.acro %>% left_join(site.therm) %>% 
  rename(Ind = Vial) %>% 
  dplyr::select(-Depth)

```



```{r}
pred2 <- env[4:22]
```

```{r}
pred.pca2 <- rda(pred2, scale = T)
summary(pred.pca2)$cont
```

```{r}
screeplot(pred.pca2, main = "Screeplot: Eigenvalues of Wolf Predictor Variables")
```

```{r}
round(scores(pred.pca2, choices=1:10, display="species", scaling=0), digits=3)
```


```{r}
library(corrplot)
pred2 %>% cor %>% corrplot()

pred3 <- pred2 %>% dplyr::select(recent.meanDHW, DHW4, returnDHW6, MMM, catBleaching)
pred3 %>% cor %>% corrplot()

```


##Run backward stepwise model
```{r}
library(vegan)
# Use backward selection to identify significant thermal environmental variables
ord_acro_therm <- dbrda(gen.imp ~ ., data = pred3)

sort(vif.cca(ord_acro_therm)) 

os_acro_therm <- ordistep(ord_acro_therm, direction = "forward", permutations = 999)
anova(os_acro_therm, by = 'margin')

```






#left join thermal and environmental data. 
```{r}
therm_red <- therm %>% dplyr::select(Ind, recent.meanDHW, DHW4, returnDHW6, MMM, catBleaching)

env <- therm_red %>% left_join(., all.var) %>% 
  dplyr::select(-Ind) %>% 
  rename(AHCL1 = "Cluster 1") %>% 
   rename(AHCL2 = "Cluster 2") %>% 
   rename(AHCL3 = "Cluster 3")
```

#run partial RDA
```{r}
acro.rda.therm <- rda(gen.imp ~ recent.meanDHW + DHW4 + returnDHW6 + catBleaching + Condition(AHCL2 + AHCL3 + MEM1 + MEM2 + MEM3), data=env, scale=T)

acro.rda.therm
```

Cluster 2, Cluster 3, EnvPC1, EnvPC2, MEM1, MEM2, and MEM3**

```{r}
summary(acro.rda.therm)
anova(acro.rda.therm)
```


```{r}
RsquareAdj(acro.rda.therm)
#adjusted R squared does not explain a lot. we need to use conditional factors to explain population structure before running the thermal models. 
```

```{r}
vif.cca(acro.rda.therm)
```

```{r}
plot(acro.rda.therm, scaling=3)  ## d.fault is axes 1 and 2
```
**SNPs are in red and the individuals are the black circles. Blue vectors are enviro predictors** 


```{r}
load.rda <- summary(acro.rda.therm)$species[,1:3]

hist(load.rda[,1], main="Loadings on RDA1")
hist(load.rda[,2], main="Loadings on RDA2")
hist(load.rda[,3], main="Loadings on RDA3") 
```
**Here we are looking at normality at the first 3 PCs**


```{r}
outliers <- function(x,z){
  lims <- mean(x) + c(-1, 1) * z * sd(x) ## f.nd loadings +/- z SD from mean loading     
  x[x < lims[1] | x > lims[2]]           # locus names in these tails
}
```
**This function just identifies SNPs that load in the tails of the distributions from the 3 histograms above** 



```{r}
cand1 <- outliers(load.rda[,1], 3) ## 20 SNPs
cand2 <- outliers(load.rda[,2], 3) ## 16 SNPs
cand3 <- outliers(load.rda[,3], 3) ## 22 SNPs

acro.rda.cand <- c(names(cand1), names(cand2), names(cand3)) ## j.st the names of the candidates

length(acro.rda.cand[duplicated(acro.rda.cand)]) ## 2 duplicate detections (detected on multiple RDA axes)
```
**We are using a stringent cut-off of 3 SD outside of normal distribution. We can alter this but 3 is a good standard to keep loci under very strong selection and we still get 56 SNPs** 

```{r}
#remove SNPs that are duplicated on multiple PCs
acro.rda.cand <- acro.rda.cand[!duplicated(acro.rda.cand)] ## 1.4 unique candidates 
```


```{r}
# Set up the color scheme for plotting:
bgcol  <- ifelse(colnames(gen) %in% acro.rda.cand, 'gray32', '#00000000')
snpcol <- ifelse(colnames(gen) %in% acro.rda.cand, 'red', '#00000000')

## a.es 1 & 2 - zooming in to just the SNPs here...
plot(acro.rda.therm, type="n", scaling=3, xlim=c(-1,1), ylim=c(-1,1), main="Acropora RDA, axes 1 and 2")
points(acro.rda.therm, display="species", pch=21, cex=1, col="gray32", bg='#f1eef6', scaling=3)
points(acro.rda.therm, display="species", pch=21, cex=1, col=bgcol, bg=snpcol, scaling=3)
text(acro.rda.therm, scaling=3, display="bp", col="#0868ac", cex=1)

## a.es 2 & 3
plot(acro.rda.therm, type="n", scaling=3, xlim=c(-1,1), ylim=c(-1,1), choices=c(2,3), main="Acropora RDA, axes 2 and 3")
points(acro.rda.therm, display="species", pch=21, cex=1, col="gray32", bg='#f1eef6', scaling=3, choices=c(2,3))
points(acro.rda.therm, display="species", pch=21, cex=1, col=bgcol, bg=snpcol, scaling=3, choices=c(2,3))
text(acro.rda.therm, scaling=3, display="bp", col="#0868ac", cex=1, choices=c(2,3))
```

**Which predictors are most strongly correlated? 
```{r}
intersetcor(acro.rda.therm)[,1:3]
```
**Recent mean DHW is most strongly correlated with RDA1 and 2; return DHW6 with RDA2; DHW4 quite homogenous across all 3 axes; cat bleaching mostly with RDA3** 





##Now we can plot to see which outlier SNPs are associated with what thermal variables
```{r}

## just the names of the candidates
acro.rda.cand <- c(names(cand1), names(cand2), names(cand3)) 

ncand <- length(cand1) + length(cand2) + length(cand3)
ncand

#58 SNPs using a SD of 3 
```


```{r}
cand1 <- cbind.data.frame(rep(1,times=length(cand1)), names(cand1), unname(cand1))
cand2 <- cbind.data.frame(rep(2,times=length(cand2)), names(cand2), unname(cand2))
cand3 <- cbind.data.frame(rep(3,times=length(cand3)), names(cand3), unname(cand3))

colnames(cand1) <- colnames(cand2) <- colnames(cand3) <- c("axis","snp","loading")

cand <- rbind(cand1, cand2, cand3)
cand$snp <- as.character(cand$snp)
```


```{r}
foo <- matrix(nrow=(ncand), ncol=4)  # 8 columns for 8 predictors
colnames(foo) <- c("recent.meanDHW", "DHW4", "returnDHW6", "catBleaching")

# Initialize a matrix to store correlations
for (i in 1:length(cand$snp)) {
  nam <- cand[i,2]
  snp.gen <- gen[,nam]
  foo[i,] <- apply(env,2,function(x) cor(x,snp.gen))
}

cand <- cbind.data.frame(cand,foo)  
head(cand)
```


```{r}
length(cand$snp[duplicated(cand$snp)])  
cand <- cand[!duplicated(cand$snp),] # remove duplicate detections
```

```{r}
#function to add the main predictor for each candidate SNP, and to give the correlation. 
for (i in 1:length(cand$snp)) {
  bar <- cand[i,]
  cand[i,8] <- names(which.max(abs(bar[4:7]))) # gives the variable
  cand[i,9] <- max(abs(bar[4:11]))              # gives the correlation
}

colnames(cand)[8] <- "predictor"
colnames(cand)[9] <- "correlation"

table(cand$predictor) 
```


```{r}
#here we assign a colour to each predictor in order to colour code candidate SNPs in the plot
sel <- cand$snp
env <- cand$predictor
env[env=="MEM1"] <- '#1f78b4'
env[env=="MEM2"] <- '#a6cee3'
env[env=="MEM3"] <- '#6a3d9a'
env[env=="pop.DHW2020"] <- '#e31a1c'
env[env=="pop.DHW3"] <- '#33a02c'
env[env=="pop.DHW6"] <- '#ffff33'
env[env=="pop.maxDHW"] <- '#fb9a99'
env[env=="pop.varSST"] <- '#b2df8a'

# color by predictor:
col.pred <- rownames(acro.rda$CCA$v) # pull the SNP names

for (i in 1:length(sel)) {           # color code candidate SNPs
  foo <- match(sel[i],col.pred)
  col.pred[foo] <- env[i]
}

col.pred[regexpr('#.*', col.pred)==-1] <- '#f1eef6' # non-candidate SNPs
empty <- col.pred
empty[grep("#f1eef6",empty)] <- rgb(0,1,0, alpha=0) # transparent
empty.outline <- ifelse(empty=="#00FF0000","#00FF0000","gray32")
bg <- c('#1f78b4','#a6cee3','#6a3d9a','#e31a1c','#33a02c','#ffff33','#fb9a99','#b2df8a')
```



```{r}
# axes 1 & 2
plot(acro.rda, type="n", scaling=3, xlim=c(-1,1), ylim=c(-1,1))
points(acro.rda, display="species", pch=21, cex=1, col="gray32", bg=col.pred, scaling=3)
points(acro.rda, display="species", pch=21, cex=1, col=empty.outline, bg=empty, scaling=3)
text(acro.rda, scaling=3, display="bp", col="#0868ac", cex=1)
legend("bottomright", legend=c("MEM1","MEM2","MEM3", "pop.DHW2020", "pop.DHW3", "pop.DHW6", "pop.maxDHW","pop.varSST"), bty="n", col="gray32", pch=21, cex=1, pt.bg=bg)

# axes 1 & 3
plot(acro.rda, type="n", scaling=3, xlim=c(-1,1), ylim=c(-1,1), choices=c(1,3))
points(acro.rda, display="species", pch=21, cex=1, col="gray32", bg=col.pred, scaling=3, choices=c(1,3))
points(acro.rda, display="species", pch=21, cex=1, col=empty.outline, bg=empty, scaling=3, choices=c(1,3))
text(acro.rda, scaling=3, display="bp", col="#0868ac", cex=1, choices=c(1,3))
legend("bottomright", legend=c("MEM1","MEM2","MEM3", "pop.DHW2020", "pop.DHW3", "pop.DHW6", "pop.maxDHW","pop.varSST"), bty="n", col="gray32", pch=21, cex=1, pt.bg=bg)

```


